refnx -- Neutron and X-ray reflectometry analysis in Python

### 2018-07-24 paper/manuscript.tex
Use en dash in title
Further latex/cls fixes

\begin{document}
\title{refnx - Neutron and X-ray reflectometry analysis in Python}
\cauthor[a]{Andrew~R. J.}{Nelson}{andrew.nelson@ansto.gov.au}{}
\author[b]{Stuart~W.}{Prescott}

@@

\section{Introduction}
The use of specular X-ray and neutron reflectometry for morphological characterisation of thin films on the approximate size range 10 to \SI{5000}{\angstrom} has exploded over the past years \cite{Wood2017, Daillant2009}. Most neutron and X-ray sources have instruments to perform reflectometry measurements, and there is an ongoing need for accessible software programs for users of those instruments to analyse their data in a straightforward fashion, including the co-refinement of multiple contrast datasets.  Several programs are available for this purpose, with a variety of different features \cite{Bjorck2007, Gerelli2016, Kienzle2011, Nelson2006}. These programs typically create a model of the interface, and incrementally refine the model against the data using least-squares, or use Bayesian approaches \cite{Sivia2006, 2010arXiv1008.4686H} to examine the posterior probability distribution of the parameters (i.e. the statistical variation of the parameters in a model).

@@

Here we outline a new reflectometry analysis package, \refnx\ (version number 0.1 is used in this paper \cite{refnx}), that can address the reproducibility issue for the reflectometry community as it is specifically designed for use in \Jupyter\ notebooks \cite{Kluyver:2016aa} using a Python kernel \footnote{We do not mean that other programs are irreproducible, rather that the information provided by scientists in journal articles is often lacking.}.
These notebooks are a mix of executable code cells and markup. An analysis performed in such a notebook can be appended as supporting information, along with the data, to an article. Readers can then replicate the exact data analysis, provided they have set up the same computing environment. Setting up the computing environment is simplified using the \conda\ package manager \cite{conda}, and an environment file (although other approaches are available). To that end the \Jupyter\ notebooks for the example analysis performed in this paper are appended as supporting information.

@@

The posterior distribution, $p(\theta | I, D)$, is the probability of the parameters given the data and prior knowledge about the system.  The posterior probability is sampled by encoding the likelihood and prior distributions and using Markov Chain Monte Carlo (MCMC) techniques \cite{emcee, ptemcee}. At the end of an MCMC run the parameter set possesses a number of samples (called a 'chain'), whose statistics reveal the distribution and covariance of the parameters, the spread of the model around the data, and in a reflectometry context the range of structures that are consistent with the data.

\section{Method}
\refnx\ is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter\ object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution representing the prior. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ can be applied using constraints.

The \Structure\ object represents the interfacial model, comprising of individual \Component\ objects in series. Each \Component\ represents a subset of the interface and contains several physically relevant Parameters. The simplest \Component\ is a \Slab\, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab. More sophisticated components include \LipidLeaflet\ and \Spline. \Spline\ is designed for freeform modelling of an SLD profile using spline interpolation. In addition to its constituent parameters each \Component\ can also contribute to the log-prior probability. This is useful when a \Component\ has a derived value, such as surface coverage, which is already known, and can be included as prior knowledge of the system. Each \Component\ has a \emph{slabs} property that represents the strata for its particular subset of the interface. A \Slab\ object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet\ can be made of two slices (head/tail regions), but the \Spline\ Component has many. These slices have uniform SLD, and use the Nevot-Croce approach \cite{Nevot1980} to describe roughness between them. If a different roughness model is required then a 'microsliced' \Component\ could be written to accomodate that.

@@

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting graphical user interface (GUI) can be created in a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the 'To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}. For a comprehensive demonstration of multiple contrast co-refinement see the annotated notebook in the supporting information.

@@

\subsection{Example}
Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach.

@@

Here \LipidLeaflet\ is used to co-refine three contrasts (\ce{D_2O}, \ce{Si} contrast match [hdmix, SLD=\SI{2.07E-6}{\per\square\angstrom}], and \ce{H_2O}) of a 1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC) bilayer at the solid-liquid interface, Figure \ref{fig:global_fit} \footnote{The validity of \LipidLeaflet\ does depend on the area per molecule being equal for the headgroup and tailgroup regions, as pointed out by Gerelli \cite{Gerelli2016}, which can be violated if there are guest molecules that insert in the membrane.}. Two \LipidLeaflet\ are required to describe the inner and outer leaflets of a bilayer, the component contains an attribute which can reverse the direction of one of the leaflets. The use of individual objects to describe each leaflet leads to great flexibility; it becomes easy to model asymmetric bilayers (inner leaflet can be a different lipid to the outer lipid), and one can model interstitial water layers between the leaflets as well.

The \Jupyter\ notebook used for the analysis, \emph{lipid.ipynb}, is available in the supporting information. The corner plot (Figure \ref{fig:corner}) produced from the MCMC analysis shows the covariance between parameters, with an area per molecule of \SI{57.0 \pm0.15}{\square\angstrom}. Figures~\ref{fig:global_fit} and \ref{fig:d2o_sld_spread} show the probability distribution of the generative model around the data and in the SLD profile. These are obtained by random samples from the MCMC chain. The spread in SLD profiles is used to determine what range of structures is consistent with the data. Multi-modalities in these SLD profiles can be due to statistical uncertainties, the Q ranges measured, and the loss of phase information in NR \cite{Majkrzak1999, Heinrich2009}.
%TODO A reference for V_m?
%FIXME units of area per molecule were wrong; check

@@

\begin{figure}
  \includegraphics[width=175mm]{./supporting_information/corner.png}
  \caption{Corner plot for the varying parameters of DMPC bilayers measured at three contrasts. The sampling took $\sim$13 minutes on a 3.2GHz quad-core computer for 30 saved steps, corresponding to 6000 samples, with the steps being thinned by a factor of 100.}
  \label{fig:corner}
\end{figure}

\section{Distribution and Modification}
Each submodule in \refnx\ possesses its own unit testing code for checking that the functions and classes in the module operate correctly, both individually and collectively. For example, there are tests that check that the reflectivity of a model is calculated correctly, or that the behaviour of a function is correct for the different possible inputs and code paths through it. Since the test suite is an integral part of the package each installation is testable. In addition, there is a benchmarking suite to track changes in performance, specifically the speed of critical calculations, over time.

@@

\section{Comments on reproducibility of analyses}
In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreatable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are often omitted in the literature. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx\ in a \Jupyter\ notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx, \SciPy, \NumPy, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda\ package manager, by installing from the source at a given version tag in the git repository, or by \pip. \conda\ can use environment files to recreate a specific setup. An alternative way of recreating the environment is by using a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter\ notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here. With closed source or proprietary software the ability to return to a specific software version/environment can be difficult, different versions can have modifications that can change the output of an analysis. In addition accessibility (due to cost, etc) to the wider scientific community can also hinder reproducibility.
Moreover, there are important ramifications for verifiability. \refnx\ is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx\ is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.

\section{Conclusions}\label{conclusions}
\refnx\ is a powerful tool for least-squares or Bayesian analysis of neutron and X-ray reflectometry data that is ideally usable for reproducible research with \Jupyter\ notebooks, and with extensibility in mind. Its features include: MCMC sampling of posterior distribution for parameters, structural models constructed from modular components with physically relevant parameterisation, algebraic inter-parameter constraints, mixed area models, co-refinement of multiple datasets, probability distributions for parameter bounds used directly for log-prior terms, and a (\Jupyter\ ) \ipywidgets\ GUI.

\section{Acknowledgements}
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Andrew McCluskey (Bath University) and Isaac Gresham (UNSW) for important feedback on \refnx\ development.

\subsection{Supporting information}
\textbf{gui.ipynb} - \Jupyter\ notebook used to create the GUI screenshot.\\
\textbf{lipid.ipynb} - \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{lipid.pdf} - PDF view of the \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{refnx-paper.yml} - \conda\ environment file to reproduce the analysis environment in this paper.

>>>>

\begin{document}
\title{refnx -- Neutron and X-ray reflectometry analysis in Python}
\cauthor[a]{Andrew~R. J.}{Nelson}{andrew.nelson@ansto.gov.au}{}
\author[b]{Stuart~W.}{Prescott}

@@

\section{Introduction}

The use of specular X-ray and neutron reflectometry for morphological characterisation of thin films on the approximate size range 10 to \SI{5000}{\angstrom} has exploded over the past years \cite{Wood2017, Daillant2009}. Most neutron and X-ray sources have instruments to perform reflectometry measurements, and there is an ongoing need for accessible software programs for users of those instruments to analyse their data in a straightforward fashion, including the co-refinement of multiple contrast datasets.  Several programs are available for this purpose, with a variety of different features \cite{Bjorck2007, Gerelli2016, Kienzle2011, Nelson2006}. These programs typically create a model of the interface, and incrementally refine the model against the data using least-squares, or use Bayesian approaches \cite{Sivia2006, 2010arXiv1008.4686H} to examine the posterior probability distribution of the parameters (i.e.\ the statistical variation of the parameters in a model).

@@

Here we outline a new reflectometry analysis package, \refnx\ (version number 0.1 is used in this paper \cite{refnx}), that can address the reproducibility issue for the reflectometry community as it is specifically designed for use in \Jupyter\ notebooks \cite{Kluyver:2016aa} using a Python kernel.\footnote{We do not mean that other programs are irreproducible, rather that the information provided by scientists in journal articles is often lacking.}
These notebooks are a mix of executable code cells and markup. An analysis performed in such a notebook can be appended as supporting information, along with the data, to an article. Readers can then replicate the exact data analysis, provided they have set up the same computing environment. Setting up the computing environment is simplified using the \conda\ package manager \cite{conda}, and an environment file (although other approaches are available). To that end the \Jupyter\ notebooks for the example analysis performed in this paper are appended as supporting information.

@@

The posterior distribution, $p(\theta | I, D)$, is the probability of the parameters given the data and prior knowledge about the system.  The posterior probability is sampled by encoding the likelihood and prior distributions and using Markov Chain Monte Carlo (MCMC) techniques \cite{emcee, ptemcee}. At the end of an MCMC run the parameter set possesses a number of samples (called a `chain'), whose statistics reveal the distribution and covariance of the parameters, the spread of the model around the data, and in a reflectometry context the range of structures that are consistent with the data.

\section{Method}

\refnx\ is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter\ object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution representing the prior. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ can be applied using constraints.

The \Structure\ object represents the interfacial model, comprising of individual \Component\ objects in series. Each \Component\ represents a subset of the interface and contains several physically relevant Parameters. The simplest \Component\ is a \Slab\, which has a uniform scattering length density (SLD), thickness, roughness, and volume fraction of solvent. The simplest models are simply a series of \Slab. More sophisticated components include \LipidLeaflet\ and \Spline. \Spline\ is designed for freeform modelling of an SLD profile using spline interpolation. In addition to its constituent parameters each \Component\ can also contribute to the log-prior probability. This is useful when a \Component\ has a derived value, such as surface coverage, which is already known, and can be included as prior knowledge of the system. Each \Component\ has a \emph{slabs} property that represents the strata for its particular subset of the interface. A \Slab\ object has a single slice because it is a single thickness of uniform SLD. A \LipidLeaflet\ can be made of two slices (head/tail regions), but the \Spline\ Component has many. These slices have uniform SLD, and use the Nevot-Croce approach \cite{Nevot1980} to describe roughness between them. If a different roughness model is required then a `microsliced' \Component\ could be written to accomodate that.

@@

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting graphical user interface (GUI) can be created in a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the `To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}. For a comprehensive demonstration of multiple contrast co-refinement see the annotated notebook in the supporting information.

@@

\subsection{Example}

Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach.

@@

Here \LipidLeaflet\ is used to co-refine three contrasts (\ce{D_2O}, \ce{Si} contrast match [hdmix, SLD=\SI{2.07E-6}{\per\square\angstrom}], and \ce{H_2O}) of a 1,2-dimyristoyl-sn-glycero-3-phospho\-choline (DMPC) bilayer at the solid-liquid interface, Figure~\ref{fig:global_fit}.\footnote{The validity of \LipidLeaflet\ does depend on the area per molecule being equal for the headgroup and tailgroup regions, as pointed out by Gerelli \cite{Gerelli2016}, which can be violated if there are guest molecules that insert in the membrane.} Two \LipidLeaflet\ are required to describe the inner and outer leaflets of a bilayer, the component contains an attribute which can reverse the direction of one of the leaflets. The use of individual objects to describe each leaflet leads to great flexibility; it becomes easy to model asymmetric bilayers (inner leaflet can be a different lipid to the outer lipid), and one can model interstitial water layers between the leaflets as well.

The \Jupyter\ notebook used for the analysis, \emph{lipid.ipynb}, is available in the supporting information. The corner plot (Figure~\ref{fig:corner}) produced from the MCMC analysis shows the covariance between parameters, with an area per molecule of \SI{57.0 \pm0.15}{\square\angstrom}. Figures~\ref{fig:global_fit} and \ref{fig:d2o_sld_spread} show the probability distribution of the generative model around the data and in the SLD profile. These are obtained by random samples from the MCMC chain. The spread in SLD profiles is used to determine what range of structures is consistent with the data. Multi-modalities in these SLD profiles can be due to statistical uncertainties, the Q ranges measured, and the loss of phase information in NR \cite{Majkrzak1999, Heinrich2009}.
%TODO A reference for V_m?
%FIXME units of area per molecule were wrong; check

@@

\begin{figure}
  \includegraphics[width=175mm]{./supporting_information/corner.png}
  \caption{Corner plot for the varying parameters of DMPC bilayers measured at three contrasts. The sampling took $\sim$13 minutes on a \SI{3.2}{GHz} quad-core computer for 30 saved steps, corresponding to 6000 samples, with the steps being thinned by a factor of 100.}
  \label{fig:corner}
\end{figure}

\section{Distribution and Modification}

Each submodule in \refnx\ possesses its own unit testing code for checking that the functions and classes in the module operate correctly, both individually and collectively. For example, there are tests that check that the reflectivity of a model is calculated correctly, or that the behaviour of a function is correct for the different possible inputs and code paths through it. Since the test suite is an integral part of the package each installation is testable. In addition, there is a benchmarking suite to track changes in performance, specifically the speed of critical calculations, over time.

@@

\section{Comments on reproducibility of analyses}

In order for a given scattering analysis to be fully reproducible by others a general set of conditions need to be met:
\begin{itemize}
  \item the datasets used need to be deposited with a journal article, or be freely available.
  \item the exact software environment needs to be recreatable.
  \item the exact ordered set of steps taken during the analysis needs to be listed.
\end{itemize} 
Each of these points are often omitted in the literature. For example, the use of different software versions may change the output of an analysis, or the use of a GUI program may not record the full set of steps, or options, applied by a user.
Whilst it is unable to meet the first criteria by itself, the use of \refnx\ in a \Jupyter\ notebook can fulfil the other two requirements, providing a little care is taken. The exact software environment can be recreated by noting down the versions of the software packages used during an analysis (\refnx, \SciPy, \NumPy, Python, etc). At a later date those exact versions can be installed in the same Python version using one of: the \conda\ package manager, by installing from the source at a given version tag in the git repository, or by \pip. \conda\ can use environment files to recreate a specific setup. An alternative way of recreating the environment is by using a virtual machine, such as Docker. The ordered set of steps to perform the analysis is recordable in the \Jupyter\ notebook.
The usefulness of open-source software in a git (or other version controlled) repository must be emphasised here. With closed source or proprietary software the ability to return to a specific software version/environment can be difficult, different versions can have modifications that can change the output of an analysis. In addition accessibility (due to cost, etc) to the wider scientific community can also hinder reproducibility.
Moreover, there are important ramifications for verifiability. \refnx\ is based on a fully open software stack, with good unit test coverage. The user can run tests for each component and inspect parts for correctness. For example, the behaviour of the reflectivity calculation in \refnx\ is checked from first principles in the test suite; and can be done now and in several years time. If problems are discovered, they can be corrected. With a closed source program such checking is harder, one has to examine the inputs/outputs of a black box without knowing what is happening in between, making it harder to find errors; one has to rely on trust. Whilst this may not be an issue for relatively simple calculations, it may be for complex algorithms, such as those involved with minimisation.

\section{Conclusions}\label{conclusions}

\refnx\ is a powerful tool for least-squares or Bayesian analysis of neutron and X-ray reflectometry data that is ideally usable for reproducible research with \Jupyter\ notebooks, and with extensibility in mind. Its features include: MCMC sampling of posterior distribution for parameters, structural models constructed from modular components with physically relevant parameterisation, algebraic inter-parameter constraints, mixed area models, co-refinement of multiple datasets, probability distributions for parameter bounds used directly for log-prior terms, and a (\Jupyter\ ) \ipywidgets\ GUI.

\ack{Acknowledgements:
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Andrew McCluskey (Bath University) and Isaac Gresham (UNSW) for important feedback on \refnx\ development.}

\section{Supporting information}

\noindent
\textbf{gui.ipynb} - \Jupyter\ notebook used to create the GUI screenshot.\\
\textbf{lipid.ipynb} - \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{lipid.pdf} - PDF view of the \Jupyter\ notebook used for the lipid analysis example.\\
\textbf{refnx-paper.yml} - \conda\ environment file to reproduce the analysis environment in this paper.
